---
title: 'Bios 301: Assignment 3'
author: Utsav Kumar
output: pdf_document
---
  
  *Due Tuesday, 11 November, 1:00 PM*
  
  $5^{n=day}$ points taken off for each day late.

50 points total.

Submit a single knitr file (named `homework3.rmd`), along with a valid PDF output file. Inside the file, clearly indicate which parts of your responses go with which problems (you may use the original homework document as a template). Add your name as `author` to the file's metadata section. Raw R code/output or word processor files are not acceptable.

### Question 1 ###

**15 points**

**20 points**

The game of craps is played as follows. First, you roll two six-sided dice; let x be the sum of the dice on the first roll. If x = 7 or 11 you win, otherwise you keep rolling until either you get x again, in which case you also win, or until you get a 7 or 11, in which case you lose.

Write a program to simulate a game of craps. You can use the following snippet of code to simulate the roll of two (fair) dice:

```{r}
x <- sum(ceiling(6*runif(2)))
```

1. The instructor should be able to easily import and run your program (function), and obtain output that clearly shows how the game progressed.  Set the RNG seed with `set.seed(100)` and show the output of three games. (15 points)

    ```{r}
    set.seed(100)
    game_craps <- function (print.out)
    # set print.out as 0 if don't want to print values for each dice roll
    {
      counter <- 1
      # If the player wins then games.won is changed to else kept 0
      games.won <- 0
      moves = c()
      x <-sum(ceiling(6*runif(2)))
      x0 <- x
      moves[counter] = x0
      if (print.out != 0)
      {cat ("For dice roll ",counter , " the value is ", moves[counter], "\n")}
      if (x0 == 7 || x0 == 11)
      {
        if (print.out != 0)
        {cat("you win in your first move \n \n" )}
        counter = counter + 1
        games.won <- games.won + 1
      }
      else
      {
        while (1)
        {
          x <- sum(ceiling(6*runif(2)))
          counter <- counter +1
          moves[counter] = x
          if (print.out != 0)
          {cat ("For dice roll ",counter , " the value is ", moves[counter], "\n")}
          if (x == x0) 
          {
            if (print.out != 0)
            {cat ("you win \n \n")}
            games.won <- games.won + 1
            break
          }
          if (x == 7 || x == 11)
          {
            if (print.out != 0)
            {cat ("you loose \n \n")}
            break
          }
        }
      }
      return (games.won)
    }
    
    # Results for first 10 games for the set.seed = 100
    play <- c()
    no_games <- 3
    for (i in 1:no_games)
    {
      play[i] <- game_craps(1) 
    }

    ```


1. Find a seed that will win ten straight games.  Consider adding an argument to your function that disables output. Show the output of the ten games. (5 points)

    ```{r}

    # Using the same function as above mentioned and trying to find the seed value 
    # for which first ten plays results in win
    seed.value <- 0
    # stores the seed value for which total games won = 10
    for (i in 1:1000)
    # Loop to find the seed value running till 1000 
    {
      total.games.won <- 0
      # stroring total games won for the first 10 play
      set.seed(i)
      for (j in 1:10)
      # First ten games are played for selected seed value
      {
        games.won <- game_craps (0)
        # returns 1 if game is won
        total.games.won = total.games.won + games.won
        # counting total games won out of 10
      }
      if (total.games.won == 10)
      # if all games for the first 10 games are won
      {
        seed.value = i
        break
      }
    }
    
    cat ("Value of the seed for which first 10 straight games are won =", seed.value)
    set.seed (seed.value)
    play.10straight.win.game <- c ()
    # setting up the seed for which all first 10 games are won
    for (i in 1:10)
    {
      # printing the game proceedings so print.out is 1
      play.10straight.win.game[i] <- game_craps (1)
    }
    
    ```

### Question 2 ###

**20 points**

Code a function that does golden section search, and use this function to find all of the global maxima on the following function:

$$f(x) = \left\{
\begin{array}{ll}0 & \text{if } x=0 \\
|x| \log\left(\frac{|x|}{2}\right)e^{-|x|} & \text{otherwise}
\end{array}
\right.
$$

on the interval [-10, 10].

To get an idea of what the function looks like, it might be helpful to plot it.  
  
```{r}
    # Defining function for which all maxima has to be found in the range [-10,10]
    fx <- function (x)
    {
      if (x == 0)
      {
        0
      }
      else 
      {
        abs(x)*log((abs(x)/2))*exp(-abs(x))
      }
    }

    # Golden section search method to find maxima 
    golden.section.search = function (f, minx, maxx, tolerance)
    {
       golden.ratio = 2/(sqrt(5) + 1)
       minx0 <- minx
       maxx0 <- maxx
       # Initial points for iteration
       x1 = maxx - golden.ratio*(maxx - minx)
       x2 = minx + golden.ratio*(maxx - minx)
    
       f1 = f(x1)
       f2 = f(x2)
    
       iteration = 0
    
       while (abs(maxx - minx) > tolerance)
       {
          iteration = iteration + 1
    
          if (f1 > f2)
          # If f1 > f2 it means maxima is between minx and x2 
          {
             # Set the new maxx
             maxx = x2
             # Getting the updated iteration points and respective function values
             x2 = x1
             f2 = f1
             x1 = maxx - golden.ratio*(maxx - minx)
             f1 = f(x1)
          } 
          else 
          {
             # else maximum is between x1 and maxx 
             # setting up the new minx and maxx
             minx = x1
             # Getting the updated iteration points and respective function values
             x1 = x2
             f1 = f2
             x2 = minx + golden.ratio*(maxx - minx)
             f2 = f(x2)
          }
       }
    
       # Using the middle point of the final minx and maxx for the answer
       cat('total number of iterations = ', iteration, '\n')
       cat('final minx =', minx, '\n')
       cat('final maxx =', maxx, '\n')
       finalx = (minx + maxx)/2
       cat('point of maxima for function f for the range [',minx0,',',maxx0,' ] 
           exists at x =', finalx, '\n')
    }
    
    # Since the function is even and plotting the graphs shows maxima exists
    # on either side of origin

        golden.section.search (fx, -10, 0, 5e-8)
        golden.section.search (fx, 0, 10, 5e-8)

```


### Question 3 ###

**10 points**

Obtain the code for using Newton's Method to estimate logistic regression parameters (`logistic.r`) and modify it to predict `death` from `weight`, `hemoglobin` and `cd4baseline` in the HAART dataset. Use complete cases only. Report the estimates for each parameter, including the intercept.

Note: The original script `logistic_debug.r` is in the exercises folder.  It needs modification, specifically, the logistic function should be defined:

```{r}
logistic <- function(x) 1 / (1 + exp(-x))
```

```{r}

    # Reading data 
    data <- read.table("~/Documents/BIOS301/Bios301/datasets/haart.csv", sep=",", head=T)
    
    # Logistic function
    logistic <- function(x) 1 / (1 + exp(-x))
    
    # extracting the required columns 
    x1 <- data[,c("cd4baseline","weight","hemoglobin")]
    y1 <- data[,c("death")]
    
    # finding dimensions
    n <- dim(x1)[1]
    k <- dim(x1)[2]
    
    x1 <- as.matrix(x1)
    y1 <- as.matrix(y1)

    # removing those datasets which are not complete
    x <- na.omit(x1)
    # defining y to the column size of x
    y <- rep(1,nrow(x))
    
    estimate_logistic <- function(x, y, MAX_ITER=10) 
    {

      n <- dim(x)[1]
      k <- dim(x)[2]
  
      x <- as.matrix(cbind(rep(1, n), x))
      y <- as.matrix(y)
  
      # Initialize fitting parameters
      theta <- rep(0, k+1)
  
      J <- rep(0, MAX_ITER)
  
      for (i in 1:MAX_ITER) 
      {

        # Calculate linear predictor
        z <- x %*% theta
        # Apply logit function
        h <- logistic(z)

        # Calculate gradient
        grad <- t((1/n)*x) %*% as.matrix(h - y)
        # Calculate Hessian
        H <- t((1/n)*x) %*% diag(array(h)) %*% diag(array(1-h)) %*% x

        # Calculate log likelihood
        J[i] <- (1/n) %*% sum(-y * log(h) - (1-y) * log(1-h))

        # Newton's method
        theta <- theta - solve(H) %*% grad
      }

      return(theta)
  }

  estimate_logistic(x, y)

```

### Question 4 ###

**5 bonus points**

Import the `addr.txt` file from the GitHub repository.  This file contains a listing of names and addresses (thanks google).  Parse each line to create a data.frame with the following columns: lastname, firstname, streetno, streetname, city, state, zip.  Keep middle initials or abbreviated names in the firstname column.  Print out the entire data.frame.

```{r eval = FALSE}

   nc<-max(count.fields("~/Documents/BIOS301/Bios301/datasets/addr.txt", sep=""))
  x<-read.table("~/Documents/BIOS301/Bios301/datasets/addr.txt", sep="",
                col.names=paste("v",1:nc,sep="."),fill=T, )
  x.df <- data.frame(x)
  print (x.df)

```

